Matching Bullets

### 2015-12-12 imaging-paper.Rnw
figures moved around a bit, but looks better now

\usepackage[USenglish]{babel}
\usepackage[protrusion=true,expansion=true]{microtype} % Better typography
\usepackage{amsmath,amsfonts,amsthm} % Math packages
\usepackage[svgnames]{xcolor} % Enabling colors by their 'svgnames'
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{fix-cm}	 % Custom font sizes - used for the initial letter in the document
\usepackage{float}
\usepackage{afterpage}

@@

    \renewcommand{\topfraction}{0.99}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.99}	% max fraction of floats at bottom

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, linetype=bullet)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between crosscuts (c).}
\end{figure}
\afterpage{\clearpage}

\subsection{Varying smoothing factor}
As mentioned, the algorithm for peak/valley detection depends on the selection of a smoothing window, called the smoothing factor. In particular, for detecting and removing the grooves prior to fitting a Loess regression, and for detecting the peaks/valleys of the loess residuals, we selected a smoothing factor of 35. We can investigate the performance of peak/valley detection across different smoothing factors. Figure \ref{fig:varysmooth} displays the resulting peaks and valleys across smoothing factors of 5, 15, and 35 respectively. It is immediately clear that a small smoothing factor like 5 is a poor choice. This results in a significant amount of noise in the data such that even just a point or two can skew the rolling average enough for a peak or valley to be detected. Given that the striation patterns are typically much larger, we are in effect muddying the waters by performing such minimal smoothing.

@@

  ggplot() + theme_bw() +
    geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), ymin=-6, ymax=6, 
              data=peaks,  alpha=0.2) +
    geom_vline(aes(xintercept=extrema, colour=factor(type)), data= peaks) +
    scale_colour_brewer(palette="Set2") + 
    scale_fill_brewer(palette="Set2") +
    theme(legend.position="none") + 
    facet_grid(smoothfactor~., labeller="label_both") +
    geom_line(aes(x=y, y=resid, group=x), data=br111.loess$data,  
              colour="grey50") +
    geom_line(aes(x=y, y=smoothed), data=smooths) +
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) 
@
%'     }
%' \end{subfigure}    
%' \begin{subfigure}[t]{\linewidth}
%'     \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 5.\label{fig:smoothfac2}}{%
%' <<smoothfac2, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
%' get_peaks(br111.loess$data, smoothfactor = 5)$plot
%' @
%'     }
%' \end{subfigure}
%' \begin{subfigure}[t]{\linewidth}
%'     \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 65.\label{fig:smoothfac3}}{%
%' <<smoothfac3, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
%' get_peaks(br111.loess$data, smoothfactor = 65)$plot
%' @
%'     }
%' \end{subfigure}
\caption{\label{fig:varysmooth} Peak/valley detection at smoothing factors of 35, 5, and 65 respectively. Note that a smoothing factor of 5 yields enough noise that many very minimal overlapping peaks and valleys are detected, while a smoothing factor of 65 over-smoothes and causes the peaks/valleys to either disappear or be horizontally shifted from their position in the original loess residuals.}
\end{figure}

@@

\hh{XXX next step: how does this matching work in known non-matches? distribution of CMS under a null of no match? }

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 36 pairs of lands with at least sixteen CMS in common are matches. There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$cms[!bstats$match])} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$cms[!bstats$match])}. Known matches share on average \Sexpr{mean(bstats$cms[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$cms[bstats$match])}. Secondly, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by Miller~\cite{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch, between two lands that share twelve consecutively matched striae. Visually we can easily tell these two lands do not match well.
%
\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{Distribution of maximal CMS.}
<<cms-bars, echo=FALSE, fig.width=7, fig.height=3.5, out.width='\\linewidth'>>=
ggplot(data=bstats) + 
  geom_bar(aes(x=factor(cms))) + theme_bw() +
  theme(legend.position="bottom") + 
  xlab("maximum CMS")
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{All land-to-land comparisons a maximum number of CMS of thirteen or more are matches.}
<<cms-spines, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='\\linewidth'>>=
ggplot(data=bstats) + 
  geom_bar(aes(x=factor(cms), fill=match), position="fill") +
  theme(legend.position="bottom") + 
  scale_fill_brewer(palette="Paired") +
  xlab("maximum CMS")
@
\end{subfigure}
\caption{\label{fig:cms}Overview of the maximal number of consecutive maximal striae compared to matches.}
\end{figure}

>>>>

\usepackage[USenglish]{babel}
\usepackage[protrusion=true,expansion=true]{microtype} % Better typography
\usepackage{amsmath,amsfonts,amsthm} % Math packages
\usepackage[svgnames]{xcolor} % Enabling colors by their 'svgnames'
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{fix-cm}	 % Custom font sizes - used for the initial letter in the document
\usepackage{float}
%\usepackage{afterpage}

@@

    \renewcommand{\topfraction}{0.99}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.99}	% max fraction of floats at bottom
% parameters for text pages
\setcounter{topnumber}{2}
    \setcounter{bottomnumber}{2}
    \setcounter{totalnumber}{4}     % 2 may work better
    \setcounter{dbltopnumber}{2}    % for 2-column pages
    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
    %   Parameters for FLOAT pages (not text pages):
    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
	% N.B.: floatpagefraction MUST be less than topfraction !!
    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bullet, linetype=bullet)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between crosscuts (c).}
\end{figure}
%\afterpage{\clearpage}

\subsection{Varying smoothing factor}
As mentioned earlier, the algorithm for peak/valley detection depends on the selection of a smoothing window, called the smoothing factor. In particular, for detecting and removing the grooves prior to fitting a Loess regression we selected a smoothing factor of 35, while for detecting the peaks/valleys of the loess residuals a smoothing factor of 15 seems more appropriate. We can investigate the performance of peak/valley detection across different smoothing factors. Figure~\ref{fig:varysmooth} displays the resulting peaks and valleys across smoothing factors of 5, 15, and 35, respectively. The dark line corresponds to the smoothed values, while the grey line in the back shows the unsmoothed values of the surface measurements. The choice of smoothing factor is a classical decision of a bias/variance trade-off. It is immediately clear that a small smoothing factor like 5 is a poor choice. It results in a significant amount of noise in the data such that even just a point or two can skew the rolling average enough for a peak or valley to be detected. Given that the striation patterns are typically much larger, we are in effect muddying the waters by performing such minimal smoothing.

@@

  ggplot() + theme_bw() +
    geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), ymin=-6, ymax=6, 
              data=peaks,  alpha=0.2) +
    geom_vline(aes(xintercept=extrema, colour=factor(type)), data= peaks) +
    scale_colour_brewer(palette="Set2") + 
    scale_fill_brewer(palette="Set2") +
    theme(legend.position="none") + 
    facet_grid(smoothfactor~., labeller="label_both") +
    geom_line(aes(x=y, y=resid, group=x), data=br111.loess$data,  
              colour="grey50") +
    geom_line(aes(x=y, y=smoothed), data=smooths) +
  ylab(expression(paste("Crosscut values (in ",mu,"m)", sep=""))) 
@
%'     }
%' \end{subfigure}    
%' \begin{subfigure}[t]{\linewidth}
%'     \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 5.\label{fig:smoothfac2}}{%
%' <<smoothfac2, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
%' get_peaks(br111.loess$data, smoothfactor = 5)$plot
%' @
%'     }
%' \end{subfigure}
%' \begin{subfigure}[t]{\linewidth}
%'     \caption{Peak and valley detection for bullet 1-5 at a smoothing factor of 65.\label{fig:smoothfac3}}{%
%' <<smoothfac3, echo=FALSE, fig.width=7, fig.height=3, out.width='\\textwidth', warning=FALSE>>=
%' get_peaks(br111.loess$data, smoothfactor = 65)$plot
%' @
%'     }
%' \end{subfigure}
\caption{\label{fig:varysmooth} Peak/valley detection at smoothing factors of 5, 15, and 35, respectively. Note that a smoothing factor of 5 yields enough noise that many very minimal overlapping peaks and valleys are detected, while a smoothing factor of 35 might over-smoothe and cause the peaks/valleys to either disappear or be horizontally shifted from their position in the original loess residuals.}
\end{figure}

@@

In order to get a better understanding of  how  this matching works in known matches and non-matches, we next investigate the algorithm's performance in a test scenario.

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 36 pairs of lands with at least sixteen CMS in common are matches. There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$cms[!bstats$match])} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$cms[!bstats$match])}. Known matches share on average \Sexpr{mean(bstats$cms[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$cms[bstats$match])}. Secondly, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by Miller~\cite{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch, between two lands that share twelve consecutively matched striae. Visually we can easily tell these two lands do not match well.
%
\begin{figure}[hbtp]
\begin{subfigure}[t]{\linewidth}
\caption{Distribution of maximal CMS.}
<<cms-bars, echo=FALSE, fig.width=7, fig.height=3.5, out.width='\\linewidth'>>=
ggplot(data=bstats) + 
  geom_bar(aes(x=factor(cms))) + theme_bw() +
  theme(legend.position="bottom") + 
  xlab("maximum CMS")
@
\end{subfigure}
\begin{subfigure}[t]{\linewidth}
\caption{All land-to-land comparisons with a maximum number of CMS of thirteen or higher are matches.}
<<cms-spines, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='\\linewidth'>>=
ggplot(data=bstats) + 
  geom_bar(aes(x=factor(cms), fill=match), position="fill") +
  theme(legend.position="bottom") + 
  scale_fill_brewer(palette="Paired") +
  xlab("maximum CMS")
@
\end{subfigure}
\caption{\label{fig:cms}Overview of the maximal number of consecutive maximal striae compared to matches.}
\end{figure}
